NumPy---Array Programming in Python

### 2020-02-06 body.tex
Redraw fig. 3: protocol. Still need to redraw icons.
Minor edits

NumPy operates on in-memory arrays using the CPU. To utilize modern,
specialized storage and hardware, there has been a recent
proliferation of Python array packages. Unlike with the Numarray and
Numeric divide, it is now much harder for these new libraries to
fracture the user community---given how much work already builds
on-top of NumPy.  However, to provide the ecosystem with access to
new and exploratory technologies, NumPy is transitioning into a
central coordinating mechanism that specifies a well-defined array
programming \emph{Application Programming Interface} (API) and
dispatches it, as appropriate, to specialized array implementations.

@@

\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights some of the essential tools and libraries, which depend
   on NumPy.  Since most of the ecosystem
   builds on NumPy arrays and its rich and powerful collection of array-aware
   functionality, the simplest way for other array implementations to become
   available to the vast scientific Python ecosystem is by using the NumPy API and array
   protocols (Fig.~\ref{fig:array-protocol}).
  }
  \label{fig:ecosystem}
\end{figure}

@@

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted the usage of
scientific Python.  Examples of its impactful application, such as the
above, now exist in almost every discipline in the natural and social
sciences.  These tools have become \emph{the
primary} software environment in many fields.  NumPy and its ecosystem are commonly
taught in university courses, boot camps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
to allow for additional performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array_protocol.jpg}

@@

NumPy was initially developed by students, faculty, and researchers to
provide an advanced, open-source array programming library for Python,
which was free to use and unencumbered by license servers, dongles, and the like.
There was a sense of building something
impactful together, for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
individuals, held a powerful attraction for many early contributors.

@@

NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs an extensive battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

>>>>

NumPy operates on in-memory arrays using the CPU. To utilize modern,
specialized storage and hardware, there has been a recent
proliferation of Python array packages. Unlike with the NumArray and
Numeric divide, it is now much harder for these new libraries to
fracture the user community---given how much work already builds
on-top of NumPy.  However, to provide the ecosystem with access to
new and exploratory technologies, NumPy is transitioning into a
central coordinating mechanism that specifies a well-defined array
programming \emph{Application Programming Interface} (API) and
dispatches it, as appropriate, to specialized array implementations.

@@

% FIXME SJW: Redraw the icons in vector format
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights some of the essential tools and libraries, which depend
   on NumPy.  Since most of the ecosystem
   builds on NumPy arrays and its rich and powerful collection of array-aware
   functionality, the simplest way for other array implementations to become
   available to the vast scientific Python ecosystem is by using the NumPy API and array
   protocols (Fig.~\ref{fig:array-protocol}).
  }
  \label{fig:ecosystem}
\end{figure}

@@

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted the usage of
scientific Python.  Examples of its significant application, such as the
above, now exist in almost every discipline in the natural and social
sciences.  These tools have become \emph{the primary}
software environment in many fields.  NumPy and its ecosystem are commonly
taught in university courses, boot camps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
to allow for additional performance optimizations.  SciPy and PyData/Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/protocol}

@@

NumPy was initially developed by students, faculty, and researchers to
provide an advanced, open-source array programming library for Python,
which was free to use and unencumbered by license servers, dongles, and the like.
There was a sense of building something consequential together,
for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
individuals, held a powerful attraction for many early contributors.

@@

NumPy is a community of practice with a \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs an extensive battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

### 2020-02-06 paper.tex
Minor edits

Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily, and, around 2000, wrote a reimplementation of
Numeric, called Numarray.
%https://lancesimms.com/programs/Python/pyraf/Docs/greenfield.pdf
Perry Greenfield created the original framework.
Many people contributed to it including Jay Todd Miller, JT Hsu, Richard
L White, Jochen Krupper, and Phil Hodge.

>>>>

Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily, and, around 2000, wrote a reimplementation of
Numeric, called NumArray.
%https://lancesimms.com/programs/Python/pyraf/Docs/greenfield.pdf
Perry Greenfield created the original framework.
Many people contributed to it including Jay Todd Miller, JT Hsu, Richard
L White, Jochen Krupper, and Phil Hodge.
